## Introduction

This report provides an analysis of the Titanic dataset, focusing on various aspects such as passenger demographics, survival rates, and the impact of different factors on survival.

## Load Required Libraries

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
```

## Data Preparation

First, we read the Titanic dataset and prepare it for analysis by creating age bins and calculating the number of family members for each passenger.

```{r}
df <- read.csv("D:\\R-statistics\\titanic.csv")

ages <- df[['Age']]
age_breaks <- c(0, 13, 19, 25, 36, 51, 66, 100)
df$age_bin <- cut(df$Age, breaks = age_breaks, right=FALSE)
df$family_members <- as.numeric(df$Parch + df$SibSp)
```

## Function to Create Pie Charts

```{r}
create_pie_chart <- function (freq_table, title){
  labels <- rownames(freq_table)
  pie(
    freq_table, 
    labels = labels,
    main = title,
    col = c("lightblue", "lightgreen", "lightcoral", "lightpink"),
    radius = 0.8,
    border = "white",
    init.angle = 90,
    cex = 0.8,
    clockwise = TRUE
  )
}
```

## Passenger Embarked Distribution

```{r}
# Number of individuals embarking on specific port
print(table(df$Embarked))
print(prop.table(table(df$Embarked)))
create_pie_chart(table(df$Embarked), "Passengers Embarked")
tbl <- table(df[!is.na(df$Embarked) & df$Embarked != "", ]$Embarked)
barplot(
  tbl, 
  main = "Passengers Embarked by Port", 
  xlab = "Port of Embarkation", 
  ylab = "Number of Passengers", 
  col = c("skyblue", "lightgreen", "salmon"),
  legend = rownames(tbl),
  ylim = c(0, 700)
)
```

## Passenger Class Distribution

```{r}
print(table(df$Pclass))
print(prop.table(table(df$Pclass)))
create_pie_chart(table(df$Pclass), "Passengers Class")
barplot(
  table(df$Pclass), 
  main = "Passengers Class Distribution", 
  xlab = "Passenger Class", 
  ylab = "Number of Passengers", 
  col = c("skyblue", "lightgreen", "salmon"), 
  names.arg = c("Class 1", "Class 2", "Class 3"),
  ylim = c(0, 500)
)
```

## Passenger Class Distribution By Gender

```{r}
gender_class_table <- table(df$Sex, df$Pclass)
print(gender_class_table)
barplot(
  gender_class_table, 
  main = "Passengers Class Distribution", 
  xlab = "Passenger Class", 
  ylab = "Number of Passengers", 
  col = c("pink", "skyblue"),
  legend = rownames(gender_class_table),
  ylim = c(0, 520),
)
```

## Passenger Gender Distribution

```{r}
print(table(df$Sex))
print(prop.table(table(df$Sex)))
create_pie_chart(table(df$Sex), "Passengers Gender")
barplot(
  table(df$Sex), 
  main = "Passengers Gender Distribution", 
  xlab = "Passenger Gender", 
  ylab = "Number of Passengers", 
  col = c("skyblue", "lightgreen", "salmon"),
  legend = rownames(table(df$Sex)),
  ylim = c(0, 600)
)
```

## Survival Distribution

```{r}
print(table(df$Survived))
print(prop.table(table(df$Survived)))
create_pie_chart(table(df$Survived), "Passengers Survival")
barplot(
  table(df$Survived), 
  main = "Passengers Survival Distribution", 
  xlab = "Survival Status", 
  ylab = "Number of Passengers", 
  col = c("lightcoral", "lightgreen"), 
  names.arg = c("Did Not Survive", "Survived"),
  ylim = c(0, 600)
)

# Create a table for survival counts based on gender
survival_gender_table <- table(df$Sex, df$Survived)

# Create a stacked bar plot
barplot(
    survival_gender_table, 
    main = "Survival Count by Gender", 
    xlab = "Survival Status", 
    ylab = "Number of Passengers", 
    col = c("lightcoral", "lightgreen"),
    legend = rownames(survival_gender_table),
    beside = FALSE,   # Make it stacked
    ylim = c(0, max(survival_gender_table) + 100),
    names.arg = c("Did Not Survive", "Survived")
)

```

## Survival Distribution by gender

```{r}
female_survival_table = table(df[df$Sex == "female",]['Survived'])
print(female_survival_table)
# relative freq
print(prop.table(female_survival_table))
male_survival_table = table(df[df$Sex == "male",]['Survived'])
print(male_survival_table)
# relative freq
print(prop.table(male_survival_table))
create_pie_chart(female_survival_table, "Female Survival")
create_pie_chart(male_survival_table, "Male Survival")
barplot(
  female_survival_table, 
  main = "Female Passengers Survival", 
  xlab = "Survival Status", 
  ylab = "Number of Female Passengers", 
  col = c("lightcoral", "lightgreen"), 
  names.arg = c("Did Not Survive", "Survived"),
  ylim = c(0, 300)
)
barplot(
  male_survival_table, 
  main = "Male Passengers Survival", 
  xlab = "Survival Status", 
  ylab = "Number of Male Passengers", 
  col = c("lightcoral", "lightgreen"), 
  names.arg = c("Did Not Survive", "Survived"),
  ylim = c(0, 500)
)

```

## Age Distribution

We create a histogram to understand the age distribution of passengers.

```{r}
hist(df$Age, breaks=10, main = "Number of Passengers by Age", xlab = "Age", ylab = "Count", border='black', col = 'lightblue', ylim=c(0, 250))

# Create a frequency table (using age range bins)
age_bins <- seq(0, 100, by = 10)  # Define age bins
age_hist <- hist(df$Age, breaks = age_bins, plot = FALSE)

# Calculate cumulative frequency
cumulative_freq <- cumsum(age_hist$counts)

# Create an ogive plot with straight lines connecting the points
plot(
    age_bins[-1], cumulative_freq, 
    type = "b",
    col = "blue", 
    lwd = 2, 
    xlab = "Age", 
    ylab = "Cumulative Frequency",
    main = "Ogive Chart for Age Distribution"
)

# Add points for better visualization
points(age_bins[-1], cumulative_freq, col = "blue", pch = 16)
```

## Age and Gender Distribution

```{r}
age_gender_table <- table(df$Sex, df$age_bin)
print(age_gender_table)
barplot(
  age_gender_table, 
  main = "Frequency of Age Ranges by Gender", 
  xlab = "Age Range", 
  ylab = "Number of Passengers", 
  col = c("lightgreen", "skyblue"),
  legend = rownames(age_gender_table),
  beside = TRUE,
  
)
```

## Survival Rates Analysis

We calculate survival rates with respect to different factors, such as the number of siblings/spouses aboard, number of parents/children aboard, gender, and age groups.

```{r}
agg_function <- function(x) {
  total = length(x)
  survived = sum(x == 1)
  not_survived = sum(x == 0)
  survival_rate = (survived / total) * 100
  return(c(Total = total, Survived = survived, Not_Survived = not_survived, Survival_Rate = round(survival_rate, 2)))
}

# Survival rate with respect to siblings/spouses
print(aggregate(Survived ~ SibSp, data = df, FUN = agg_function))
```

```{r}
# Survival rate with respect to parents/children
print(aggregate(Survived ~ Parch, data = df, FUN = agg_function))
```

```{r}
# Survival rate with respect to gender
print(aggregate(Survived ~ Sex, data = df, FUN = agg_function))
```

```{r}
# Survival rate with respect to age bins
print(aggregate(Survived ~ age_bin, data = df, FUN = agg_function))
```

## Survival rates by age

```{r}
survival_by_age_freq = table(df$Survived, df$age_bin)
print(survival_by_age_freq)
barplot(
  survival_by_age_freq, 
  main = "Survival by Age Range", 
  xlab = "Age Range", 
  ylab = "Number of Passengers", 
  col = c("lightgreen", "lightblue"),
  legend = rownames(survival_by_age_freq),
  beside = TRUE
)
```

## Survival rates by passenger class

```{r}
survival_by_pc_freq = table(df$Survived, df$Pclass)
print(survival_by_pc_freq)
barplot(
  survival_by_pc_freq, 
  main = "Survival by Passenger Class", 
  xlab = "Passenger Class", 
  ylab = "Number of Passengers", 
  col = c("lightgreen", "lightblue"),
  legend = rownames(survival_by_pc_freq),
  beside = TRUE
)
```

## Survival Rate by Age and Gender

```{r}
age_gender_survival_rate <- df %>%
  group_by(age_bin, Sex) %>%
  summarise(
    Total_Passengers = n(),
    Survived = sum(Survived == 1),
    Survival_Rate = (Survived / Total_Passengers) * 100
  )
print(age_gender_survival_rate)
ggplot(age_gender_survival_rate, aes(x = age_bin, y = Survival_Rate, color = Sex, group = Sex)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Survival Rate by Age and Gender",
       x = "Age Group",
       y = "Survival Rate (%)",
       color = "Gender")
ggplot(age_gender_survival_rate, aes(x = age_bin, y = Survival_Rate, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Survival Rate by Age Group and Gender",
    x = "Age Group",
    y = "Survival Rate (%)"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("lightcoral", "cornflowerblue"))
```

## Survival Rate by Number of Family Members

We also analyze how survival rates vary with the number of family members.

```{r}
survival_rate_wrt_family_members <- df %>%
  group_by(family_members) %>%
  summarize(
    Total_Passengers = n(),
    Survived = sum(Survived == 1),
    Survival_Rate = (Survived / Total_Passengers) * 100
  )
print(survival_rate_wrt_family_members)
ggplot(survival_rate_wrt_family_members, aes(x = family_members, y = Survival_Rate)) +
    geom_bar(stat = "identity", position = "dodge", fill = "skyblue") +  # Set the bar color
    geom_text(data = subset(survival_rate_wrt_family_members, Survival_Rate == 0),
              aes(label = "0%"), vjust = -0.5, color = "red") +  # Adding labels for 0% survival rate
    labs(title = "Survival Rate by Number of Family Members",
         x = "Number of Family Members",
         y = "Survival Rate (%)") +
    scale_x_continuous(breaks = seq(
        min(survival_rate_wrt_family_members$family_members), 
        max(survival_rate_wrt_family_members$family_members), 
        by = 1))
```

```{r}
table(df$Survived, df$Sex)
```
